---
title: Extract HVAC diagrams of all EnergyPlus example files
author: Hongyuan Jia
date: '2019-06-18'
slug: extract-hvac-diagrams
categories:
  - tech
tags:
  - eplusr
  - energyplus
  - R
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Motivation

I am wondering if in EnergyPlus, it is possible to directly connect
`Coil:Heating:Electric` and other coils with air terminal units. To achieve, I
tried to extract HVAC diagrams of all EnergyPlus example files to see if there
is any example demonstrating this.

```{r}
library(eplusr)
eplusr_option(verbose_info = FALSE)
# for converting svg to PDF
library(rsvg)
```

# Process

Get paths of all EnergyPlus v9.0 example files.

```{r}
paths_idf <- list.files(file.path(eplus_config(9.0)$dir, "ExampleFiles"),
    pattern = "\\.idf", full.names = TRUE, ignore.case = TRUE)
```

There are 584 example IDF files in total. Next step is to read all example
files.

```{r}
all_idfs <- lapply(paths_idf, read_idf)
```

Here we only care about models that contains coils and also air distribution
units.

```{r}
is_target <- vapply(all_idfs, function (idf) all(idf$is_valid_group(c("Coils", "Zone HVAC Air Loop Terminal Units"))), TRUE)
idfs <- all_idfs[is_target]
```

After filtering, there are 382 models left. Running all these models may take
years. In order to speed up, here only run every model for 1 day.

```{r}
idfs <- lapply(idfs,
    function (idf) {
        # remove original run period
        idf$RunPeriod <- NULL

        # add new run period
        idf$add(RunPeriod = list("Jan1", 1, 1, NULL, 1, 1, NULL))

        # save all models to temporary directory
        idf$save(file.path(tempdir(), basename(idf$path())), overwrite = TRUE)

        idf
    }
)
```

After saving all modified models, it is easy to run all of them using
`eplusr::run_multi()` in parallel.

```{r, results = "hide"}
jobs <- run_multi(
    model = vapply(idfs, function (idf) idf$path(), character(1)),
    weather = file.path(eplus_config(9.0)$dir, "WeatherData/USA_CO_Golden-NREL.724666_TMY3.epw"),
    NULL
)
```

Here we exclude all simulations that failed to complete and use
`eplusr:::hvac_diagram()` to call `HAVC-Diagram` post-processor to draw the
`.svg` HVAC diagrams, and in the end, call `rsvg::rsvg_pdf()` to convert `.svg`
file to `.pdf` file.

```{r, results = "hide"}
jobs[, by = index,
    {
        bnd <- paste0(tools::file_path_sans_ext(idf), ".bnd")
        if (file.exists(bnd)) {
            svg <- eplusr:::hvac_diagram(energyplus, bnd)
            rsvg::rsvg_pdf(svg, paste0(tools::file_path_sans_ext(svg), ".pdf"))
        }
    }
]
```
